GMSE: an R package for generalised management strategy evaluation

### 2017-11-09 notebook/ms/SI1.Rmd
Finish manager fitness function, move to user, and minor tweaks

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users follow identically. Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework). We first explain the key data structures used, then provide an overview of how a population of strategies is initialised, and the  subsequent processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement. We then explain the fitness functions of managers and users in more detail.

@@

The above array holds all of the information on manager and user actions. The first seven columns contain information about which entities are affected, and how they are affected. The first column `Act` identifies the type of action being performed; a value of -2 defines a direct action to a resource (e.g., culling of the resource), and a value of -1 defines direct action to a landscape (e.g., increasing yield). Positive values are currently only meaningful for `Manager_Actions`, where a value of 1 defines an action setting a uniform cost of users' direct actions on resources (i.e., costs where `Act = -2` for `User_1_Actions` and `User_2_Actions`). All other values for `Act` are meaningless in GMSE 0.3.1.9, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation). Similarly, columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in GMSE v0.3.1.9 (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`. Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the manager's utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`) of dimensions identical to `ACTION`. Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default). All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible `manager_budget` or `user_budget`, so neither can affect resource types or utilities. Columns 8-13 are also set to 10001, except where actions are allowed. Below shows the `COST` array that corresponds to the above `ACTION` array.

@@

Note that in default GMSE parameters, `culling = TRUE`, but all other actions are false. Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001. Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12). This can be verified in `ACTION` where the manager has set the cost of scaring to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users. The cost of the manager affecting the cost of user actions is always set to the `minimum_cost`; here the default `r samp_COST[3,9,1]` is used. This `minimum_cost` also defines cost values for `None`, in which the user or manager does nothing, as might occur if the manager wants to permit culling and therefore does not want to invest any of their `manager_budget` to increasing the cost of culling. Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are updated at the frequency defined.

@@

The genetic algorithm updates a single layer of the `ACTION` array, which defines to the decisions of a single agent (either the manager or a user). The corresponding layer of the `COST` array remains unchanged, and serves only to ensure that `ACTION` values do not exceed `manager_budget` or `user_budget` for managers and users, respectively. The genetic algorithm proceeds by first initialising a large population of new `ACTION` layers. In each generation, these layers crossover and mutate, generating variation in agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how much it increases the agent's utility. A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous generation is below the threshold `converge_crit` (Figure 1).

@@

**Figure 1:** Conceptual overview of the GMSE genetic algorithm

@@

At the start of each genetic algorithm, a population of size `ga_popsize` is initialised (hereafter the `POPULATION` array). This population is held in a 3D array of `ga_popsize` layers. Each layer includes an identical number of rows and columns as in `ACTION`, and one layer defines a single 'individual' in the population. The first seven columns of `ACTION` are replicated exactly for all individuals, and remain unchanged throughout the genetic algorithm thereby preserving the information about which entities are affected by actions in a given row. The remaining columns are either also replicated exactly as in `ACTION` (i.e., initialised to be the same decisions as in a previous time step), or randomly seeded with values given the constraints of `manager_budget` or `user_budget` (i.e., initialised to random decision making). The number of exact replicates initialised is set using `ga_seedrep` (if `ga_seedrep` $\geq$ `ga_popsize`, then all individuals are seeded as replicates). After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping generations begins.

@@

Once all individuals in `POPULATION` are within budget, the fitness of each individual is assessed. Fitness assessment works differently for managers versus users because managers need to consider the consequences of their decisions on user actions, and how those actions will affect resource abundance. In contrast, user actions need to consider the consequences of their decisions on resource abundance or landscape output. Fitness of an individual is defined by a real number that increases with the degree to which an individual's actions are predicted to increase their utility (recall that managers and users assign resources or landscape output a utility value). Details for how fitness is calculated are provided below.

@@

After each individual in `POPULATION` is assigned a fitness, a tournament is used to select individuals of higher fitness. Tournament selection is an especially flexible, non-parametric method that samples a subset of individuals from the total population and chooses the fittest of the subset for replacement [@Hamblin2013]. In GMSE, tournament selection proceeds by randomly sampling `ga_sampleK` individuals from the total `POPULATION` with replacement. The fitnesses of the subset of `ga_sampleK` individuals are compared, and the `ga_chooseK` individuals of highest fitness are retained (if `ga_sampleK` $\geq$ `ga_chooseK`, then all `ga_sampleK` are chosen, but this is not recommended). Tournaments selecting `ga_chooseK` individuals from random subsets of size `ga_sampleK` continue until a total of `ga_popsize` individuals are retained.

@@

Once a new set of `ga_popsize` individuals is retained through tournament selection, these individuals replace the previous `POPULATION` array. The genetic algorithm terminates if and only if a minimum number of generations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied. The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is less than `converge_crit`, then termination does not occur (it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies). If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure 1).

@@

Here we explain how candidate manager and user fitness strategies in a `POPULATION` array (see above) are calculated. We emphasise that the fitness functions used in GMSE v0.3.1.9 are intended to be heuristic tools for identifying reasonable manager and user behaviours. In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions. This practical, metaheuristic approach is consistent with the objectives of Management Strategy Evaluation [@Bunnefeld2011], and is well-suited for the use of genetic algorithms [@Hamblin2013]. @Luke2013 describes the metaheuristic approach more generally (original emphasis retained):

@@

Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3). Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above). Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate from the observation model) changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}$ is positive. To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, and the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation for the individual $i$,

@@

The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy that does not get closer to the `manage_target` abundance. Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting this value from $D_{i}$ for each individual,

$$ F_{i} = D_{max} - D_{i}. $$

We have explained how $U^{m}_{res}$ is calculated in the [above section on key data structures](#Key_data_structures_used). We now explain in more detail how individuals in the genetic algorithm calculate their actions will affect $\Delta A_{i}$. Below explains calculations for a single individual $i$ in the genetic algorithm, so we will drop the use of subscripts $i$ to refer to individuals; instead we use subscripts $j$ to refer to actions (e.g., scaring, culling, etc.).

To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ performed by users in the previous time step ($X_{j}^{old}$), and the cost of performing each action ($C_{old}^{j}$). This information is collected from `ACTION` and `COST` arrays. The individual then needs to predict how their policy (i.e., the costs that they set users to perform an action) will affect the new total number of each action performed ($X_{j}^{new}$). To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy. The predicted total number of a particular action $j$ performed is thereby calculated as,

@@

The variable $C_{new}^{j}$ defines the new cost set by the individual for action $j$. A value of 1 is added to ($X_{j}^{old}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step. Otherwise, if $X_{j}^{old} = 0$, then the manager would always assume that a change in the cost of an action would have no affect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.

The predicted consequences of $X^{new}_{j}$ for resource abundance differ for each possible action. For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `scaring = FALSE`).

@@

>>>>

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy). Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework). We first explain the key data structures used, then provide an overview of how a population of strategies is initialised, and the  subsequent processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement. We then explain the fitness functions of managers and users in more detail.

@@

The above array holds all of the information on manager and user actions. The first seven columns contain information about which entities are affected, and how they are affected. The first column `Act` identifies the type of action being performed; a value of -2 defines a direct action to a resource (e.g., culling of the resource), and a value of -1 defines direct action to a landscape (e.g., increasing yield). Positive values are currently only meaningful for `Manager_Actions`, where a value of 1 defines an action setting a uniform cost of users' direct actions on resources (i.e., costs where `Act = -2` for `User_1_Actions` and `User_2_Actions`). All other values for `Act` are meaningless in GMSE 0.3.1.9, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation). For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`.

Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.3.1.9 manager and user sub-models (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`) of dimensions identical to `ACTION`. Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default). All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible `manager_budget` or `user_budget`, so neither managers nor users can affect resource types or utilities. Columns 8-13 are also set to 10001, except where actions are allowed. Below shows the `COST` array that corresponds to the above `ACTION` array.

@@

Note that in default GMSE parameters, `culling = TRUE`, but all other actions are set to `FALSE`. Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001. Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12). This can be verified in `ACTION` where the manager has set the cost of culling to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users. The cost of the manager affecting the cost of user actions is always set to the `minimum_cost`; here the default `r samp_COST[3,9,1]` is used. This `minimum_cost` also defines cost values for `None`, in which the user or manager does nothing, as might occur if the manager wants to permit culling and therefore does not want to invest any of their `manager_budget` to increasing the cost of culling. Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are updated at the frequency defined.

@@

The genetic algorithm updates a single layer of the `ACTION` array, which defines the decisions of a single agent (either the manager or a user). The corresponding layer of the `COST` array remains unchanged, and serves only to ensure that `ACTION` values do not exceed `manager_budget` or `user_budget` for managers and users, respectively. The genetic algorithm proceeds by first initialising a large (but temporary) population of new `ACTION` layers. In each generation, these layers crossover and mutate, generating variation in potential agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how the layer is predicted to affect resources or landscape output to which the agent has assigned some utility. A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the increase in mean fitness from the previous generation is below the threshold `converge_crit` (Figure S1-1).

@@

**Figure S1-1:** Conceptual overview of the GMSE genetic algorithm

@@

At the start of each genetic algorithm, a population of size `ga_popsize` is initialised (hereafter the `POPULATION` array). This population is held in a 3D array of `ga_popsize` layers. Each layer includes an identical number of rows and columns as in `ACTION`, and one layer defines a single 'individual' in the population. The first seven columns of `ACTION` are replicated exactly for all individuals, and remain unchanged throughout the genetic algorithm thereby preserving the information about which entities are affected by actions in a given row. The remaining columns are either also replicated exactly as in `ACTION` (i.e., initialised to be the same decisions as in a previous time step), or randomly seeded with values given the constraints of `manager_budget` or `user_budget` (i.e., initialised to random decision making). The number of exact replicates initialised is set using `ga_seedrep` (if `ga_seedrep` $\geq$ `ga_popsize`, then all individuals are seeded as replicates). After the `POPULATION` of `ga_popsize` individuals is initialised, a loop simulating the adaptive evolution of `POPULATION` in non-overlapping generations begins (see Figure S1-1 above).

@@

Once all individuals in `POPULATION` are within budget, the fitness of each individual is assessed. Fitness assessment works differently for managers versus users because managers need to consider the consequences of their decisions on user actions, and how those actions will affect resource abundance. In contrast, user actions need to consider the consequences of their decisions on resource abundance or landscape output. Individual fitness is defined by a real number that increases with the degree to which an individual's actions are predicted to increase entities of positive utility and decrease entities of negative utility (recall that managers and users assign resources or landscape output a utility value). Details for how fitness is calculated are provided below.

@@

After each individual in `POPULATION` is assigned a fitness, a tournament is used to select individuals. Tournament selection is an especially flexible, non-parametric method that samples a subset of individuals from the total population and chooses the fittest of the subset for replacement [@Hamblin2013]. In GMSE, tournament selection proceeds by randomly sampling `ga_sampleK` individuals from the total `POPULATION` with replacement. The fitnesses of the subset of `ga_sampleK` individuals are compared, and the `ga_chooseK` individuals of highest fitness are retained (if `ga_sampleK` $\geq$ `ga_chooseK`, then all `ga_sampleK` are chosen, but this is not recommended). Tournaments selecting `ga_chooseK` individuals from random subsets of size `ga_sampleK` continue until a total of `ga_popsize` individuals are retained.

@@

Once a new set of `ga_popsize` individuals is retained through tournament selection, these individuals replace the previous `POPULATION` array. The genetic algorithm terminates if and only if a minimum number of generations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied. The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies). If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure S1-1).

@@

Here we explain how the fitnesses of candidate manager and user strategies in a `POPULATION` array (see above) are calculated. We emphasise that the fitness functions used in GMSE v0.3.1.9 are intended to be heuristic tools for identifying reasonable manager and user behaviours. In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions. This practical, metaheuristic approach is consistent with the objectives of Management Strategy Evaluation [@Bunnefeld2011], and is well-suited for the use of genetic algorithms [@Hamblin2013]. @Luke2013 describes the metaheuristic approach more generally (original emphasis retained):

@@

Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3). Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above). Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate from the observation model) changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive. To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, and the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,

@@

The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy is not predicted to get closer to the `manage_target` abundance. Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting $D_{i}$ from this value for each individual,

$$ F^{m}_{i} = D_{max} - D_{i}. $$

We have explained how $U^{m}_{res}$ is calculated in the [above section on key data structures](#Key_data_structures_used). We now explain in more detail how individuals in the genetic algorithm calculate how their actions will affect $\Delta A_{i}$. These calculations apply to all individuals identically, so we note that the subscript $i$ is dropped; instead, the subscript $j$ refers to actions (e.g., scaring, culling, etc.).

To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ performed by users in the previous time step ($X_{j}^{old}$), and the cost of performing each action ($C^{old}_{j}$). This information is collected from `ACTION` and `COST` arrays. The individual then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action performed ($X_{j}^{new}$). To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy. The predicted total number of a particular action $j$ performed is thereby calculated as,

@@

The variable $C^{new}_{j}$ defines the new cost set by the individual for action $j$. A value of 1 is added to ($X_{j}^{old}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step. Otherwise, if $X_{j}^{old} = 0$, then the manager would always assume that a change in the cost of an action would have no effect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.

The predicted consequences of $X^{new}_{j}$ for resource abundance differ for each possible action. For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `culling = FALSE`). For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to greatly overestimate how much the actions attempted by users will actually translate to a change in resource abundance. In practice, the default $\sigma = 0.1$ performs well. Allowed actions are predicted by managers to have the following effects:

- `scaring` is assumed to be nonlethal and therefore have an effect of zero on the total abundance of a resource.
- `culling` is assumed to have an effect of $-\sigma(1 + \lambda)$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources.
- `castration` is assumed to have an effect of $-\sigma\lambda$.
- `feeding` is assumed to have an effect of $\lambda$.
- `help_offspring` is assumed to have an effect of $\sigma$.

These effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions. The summation of $X^{new}_{j}$ for all actions defines the predicted change in resource abundance caused by the policy of an individual $i$, $\Delta A_{i}$.

@@

The previous section described the fitness function applied when individual's fitness was evaluated for managers; here we explain a separate fitness function that is applied when individuals are instead evaluated for users. Individual fitness as calculated for users ($F_{i}^{u}$) is affected by a user's utility for resources and landscape output (i.e., the predicted summed values of the landscape cells that they own), and the project change in each caused by the user's actions (i.e., the contents of their `POPULATION` layer, specifically rows 1 and 2).
